---
title: "R code for the VAR(p) model"
author: "Teresa Gual"
---

## Load the necessary libraries

```{r}
install.packages("vars")
install.packages("tseries")
install.packages("dynlm")
install.packages("forecast")
install.packages("readxl") #Para cargar el archivo Excel a RStuio
library(vars)
library(tseries)
library(forecast)
library(dynlm)
library(readxl)

```

## Load the Excel sheet with the time series data
## Specify the file path
```{r}
ruta_archivo <- "/Users/TeresaGual/Desktop/EjemploVAR-6.xlsx"
```

## Read the Excel file
```{r}
datos <- read_excel(ruta_archivo)
```

## Convert the data from each ODS to time series
```{r}
datosSDG2=ts(datos$SDG2)
datosSDG6=ts(datos$SDG6)
datosSDG7=ts(datos$SDG7)
datosSDG15=ts(datos$SDG15)
```

## Plot the time series of the 4 ODS
```{r}
ts.plot(datosSDG2,datosSDG6,datosSDG7, datosSDG15, col=c("blue","red","green", "black"), xlab = "Años desde 2000", ylab = "ODS UE")
```

## Calculate the necessary differences to achieve stationarity
```{r}
ndiffs(ldatosSDG2)
ndiffs(ldatosSDG6)
ndiffs(ldatosSDG7)
ndiffs(ldatosSDG15)
```

## Plot the transformed time series of the 4 ODS
```{r}
ts.plot(d2SDG2, d2SDG6,d2SDG7, d2SDG15, col=c("blue","red","green", "black"), xlab = "Años desde 2000", ylab = "Series ODS UE")
```

## ADF test for each series to check stationarity
```{r}
adf.test(d2SDG2)
adf.test(d2SDG6)
adf.test(d2SDG7)
adf.test(d2SDG15)
```

## Create a new object for VAR with the stationary variables
```{r}
vard2SDG2=ts(d2SDG2, start=2000)
vard2SDG6=ts(d2SDG6, start=2000)
vard2SDG7=ts(d2SDG7, start=2000)
vard2SDG15=ts(d2SDG15, start=2000)
```

## Combine the variables into a matrix
```{r}
ejvar=cbind(vard2SDG2,vard2SDG6,vard2SDG7,vard2SDG15)
```

## Select the optimal number of lags, maximum 2
```{r}
VARselect(ejvar,lag.max=4)
```

## Perform Granger causality tests to see which ODS affects which ODS
```{r}
grangertest(dSDG15~dSDG7, order=4)
```

## Study the stationarity in each column of ejvar
```{r}
adf.test(ejvar[, 1])  # Apply ADF to the first column
```

## If there is no stationarity, take second differences
```{r}
d2SDG2=diff(dSDG2)
d2SDG6=diff(dSDG6)
d2SDG7=diff(dSDG7)
d2SDG15=diff(dSDG15)
```

## Repeat the previous steps
##Create a new object for VAR with the stationary variables
```{r}
vard2SDG2=ts(d2SDG2, start=2000)
vard2SDG6=ts(d2SDG6, start=2000)
vard2SDG7=ts(d2SDG7, start=2000)
vard2SDG15=ts(d2SDG15, start=2000)

ejvar=cbind(vard2SDG2,vard2SDG6,vard2SDG7,vard2SDG15)
```

## Calculate the models for the 4 ODS
```{r}
var1<-VAR(ejvar,p=3) #p=3 if I consider 3 lags
```
